Biology background

### 2014-10-29 Rmarkdown/biology_background.Rmd
Update epigenetics section.

"Epigenetics" is a real Humpty Dumpty phrase: each author seems to believe that, "When I use a word, ... it means just what I choose it to mean - neither more nor less" [@carroll1897through]. I am sure to continue in this grand tradition.

Conrad Waddington coined the phrase in 1942 as a portmanteau of the words "epigenesis" and "genetics" (Source: [http://en.wikipedia.org/wiki/Epigenetics](http://en.wikipedia.org/wiki/Epigenetics)). He meant epigenetics as the study of how _genotypes_ give rise to phenotypes during development [@waddington1957strategy].

A popular contemporary definition of epigenetics is attributed to the epigeneticist (__CHECK SOURCE AND QUOTE__) Art Riggs: "[epigenetics is] the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" (p1 @russo1996epigenetic). This is in line with the _epi_ prefix being derived from the Greek word for "upon, near to or in addition" (__INSERT OXFORD DICTIONARY REFERENCE__), hence the idea that epigenetics encodes information "on top of" the DNA sequence. However, it is quite different to Waddington's original definition.

@@

More recently, the definition of epigenetics has taken on a "more biochemical flavour" ([@Daxinger:2010bu]). It is used to desribe marks whose heritably is yet to be proven, such as histone modifications (__SOURCE__). Sir Adrian Bird, an esteemed British geneticist, attempts to unite these definitions in the following ([@Bird:2007em]):

@@

The _epigenome_ of a cell is the set of epigenetics marks present on the cell's genome. In contrast to the genome, which is identical between cells within an individual, the epigenome is highly variable between cells within an individual. Indeed, we can identify variation in a single epigenetic mark within cells of the same cell type from the same individual (__EXAMPLE__).

Methylation, acetylation, ubiquination and phosphorylation of histones, which alter the chromatin structure, are examples of the "more biochemical flavour" of recent epigenetics. But the classic epigenetic mark is DNA methylation, which I describe in the next section and whose variability I have studied in my thesis.

- Are histone modifications epigenetic? (http://www.nature.com/news/2008/081022/full/4551023a.html)

>>>>

"Epigenetics" is a real Humpty Dumpty phrase; each author seems to believe that, "When I use [the] word, ... it means just what I choose it to mean - neither more nor less" \cite{carroll1897through}. I am sure to continue in this grand tradition.

Conrad Waddington coined the phrase in 1942 as a portmanteau of the words "epigenesis" and "genetics" \cite{Waddington:2012bu}. Waddington meant epigenetics as the study of how "processes involved in the mechanism by which the genes of the genotype bring about phenotypic effects" \cite{Waddington:2012bu}.

A popular contemporary definition of epigenetics is attributed to the epigeneticist Art Riggs that epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \cite[p1]{russo1996epigenetic}. The _epi_ prefix, derived from the Greek word for 'upon', 'near to', or 'in addition', emphasises the idea that epigenetics encodes information 'on top of' the DNA sequence. However, it is quite different to Waddington's original definition.

@@

More recently, the definition of epigenetics has taken on a "more biochemical flavour" \cite{Daxinger:2010bu} to include marks whose heritability is yet to be established. The heritability, or lack thereof, of histone modifications means that many epigeneticists do not consider these to be truly epigenetic \cite{Berger:2009ex}, and describing them as such is a sure-fire way to annoy a good percentage of your audience \cite{Ledford:2008ew}.

Sir Adrian Bird, an esteemed British geneticist and someone who is in favour of the broader definition, attempts to unite these definitions in the following ([@Bird:2007em]):

@@

The _epigenome_ of a cell is the set of epigenetic marks present on the cell's genome. In contrast to the genome, which is identical between cells within an individual, the epigenome is highly variable between cells within an individual. Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual (__TODO: cite examples or upcoming chapters__). Epigenomics is the study of the epigenome, analogous to genomics being the study of the genome. However, one can rarely study the epigenome in isolation from the underlying genetic sequence, as there is much evidence that genetic variation can drive epigenetic variation (__CITE__).

One mark that most authors agree is an epigenetic mark is DNA methylation, which I describe in the next section and whose study is the focus of my thesis.
